---
title: 'Differential expression analysis (DEA)'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How can we identify the differentially expressed proteins between experimental groups?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Conduct differential expression analysis using *limpa*
- Visualise the top differentially expressed proteins

::::::::::::::::::::::::::::::::::::::::::::::::

## Differential Expression Analysis

Now it is time to analyse our data and investigate proteins that are differentially expressed between our two experimental groups (control vs active Crohn’s disease). We will make use of the *limma* package and closely related *limpa* functions specifically designed for proteomics data. 

First, we will fit a linear model that includes both Class and Batch effects.

```{r}
# Set the control group as our reference
Class <- relevel(Class, ref = "Ctrl")
## We also need to reset our colors to match the correct labels
Class.color <- Class
levels(Class.color) <- hcl.colors(nlevels(Class), palette = "cividis")

# Fit a linear model
design <- model.matrix(~Class + Batch)
design
```

We can see both Class and Batch accounted for in our design matrix. 

Now we run the *limpa* differential expression analysis function `dpcDE()` to identify the top differentially expressed proteins in our dataset.

```{r}
fit <- dpcDE(y.protein, design, plot = TRUE)
fit <- eBayes(fit)
topTable(fit, coef = 2)
```

We can visualise the log2 expression values for differentially expressed proteins, together with standard errors.

```{r}
plotProtein(y.protein, "PRTN3_HUMAN", col = as.character(Class.color))
legend('topleft', legend = levels(Class), fill = levels(Class.color))

plotProtein(y.protein, "S10A9_HUMAN", col = as.character(Class.color))
legend('topleft', legend = levels(Class), fill = levels(Class.color))
```

## Visualise differentially expressed proteins

### Heatmap

We can filter our results to the top up to 50 most significant differentially expressed proteins, then visualise via a clustered heatmap.

```{r}
# Save our DEA results as a dataframe
results <- topTable(fit, coef = 2, number = Inf)

# Limit to the top 50 most significant results
sig_proteins <- results %>%
  filter(adj.P.Val < 0.05) %>% # filter by p value
  top_n(50, wt = abs(logFC)) %>% # filter by absolute log fold change value
  pull(Protein.Names)
expr_matrix <- y.protein$E[sig_proteins, ]

# Scale the data and visualise via heatmap
scaled_expr <- scale(t(scale(t(expr_matrix))))

pheatmap(scaled_expr,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         annotation_col = y.protein$targets,
         clustering_method = 'ward.D')
```

### Volcano plot

We can also visualise our results as a volcano plot using `EnhancedVolcano()`.

```{r}
EnhancedVolcano(results,
                lab = results$Protein.Names,
                x = 'logFC',
                y = 'adj.P.Val',
                pCutoff = 0.05,
                FCcutoff = 1.0,
                pointSize = 2.0,
                labSize = 3.0,
                drawConnectors = TRUE,
                maxoverlapsConnectors = Inf,
                lengthConnectors = unit(0.001, 'npc'),
                typeConnectors ='closed')
```


::::::::::::::::::::::::::::::::::::: keypoints 

- We can fit a linear model to our data and conduct differential expression analysis using *limpa* and *limma* functions.
- There are many different ways to visualise differentially expressed proteins.

::::::::::::::::::::::::::::::::::::::::::::::::

