---
title: 'Differential expression analysis'
teaching: 40
exercises: 5
---

:::::::::::::::::::::::::::::::::::::: questions 

- How can we identify the differentially expressed proteins between our experimental groups?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Conduct differential expression analysis using `limpa`
- Visualise the top differentially expressed proteins

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: instructor

During a live workshop, participants should work through the lesson content independently, with pauses for discussion before and after the visualisation component.

::::::::::::::::::::::::::::::::::::::::::::::::

```{r, results="hide", message=FALSE, warning = FALSE, echo=FALSE}
library(limpa)
library(readxl)
library(pheatmap)
library(EnhancedVolcano)
library(STRINGdb)
library(clusterProfiler)
library(org.Hs.eg.db)
library(arrow)
library(rpx)
library(dplyr)
library(stringr)

y.peptide <- readDIANN(file='data/MBIntroToProteomics.parquet',format="parquet",
                       q.columns = c("Q.Value","Lib.Q.Value","Lib.PG.Q.Value"),
                       q.cutoffs = 0.01)

samples_stool <- as.data.frame(readxl::read_excel('data/SampleAnnotation.xlsx'))
head(samples_stool)

# Create sample_name column that matches peptide matrix column names
samples_stool$fixed_file_name <- gsub("\\\\", "/", samples_stool$file_name)
samples_stool$fixed_file_name <- basename(samples_stool$fixed_file_name)
samples_stool$sample_name <- gsub(".mzML", "",samples_stool$fixed_file_name)

# Filter to only samples present in our dataset
samples_stool <- samples_stool %>% filter(samples_stool$sample_name %in% colnames(y.peptide$E))

# Ensure sample row order matches column order of the peptide matrix
sample_info <- samples_stool[samples_stool$sample_name %in% colnames(y.peptide$E),]
rownames(sample_info) <- sample_info$sample_name
y.peptide$E <- y.peptide$E[,rownames(sample_info)]

# Add experimental metadata to EList
y.peptide$targets <- sample_info[, c('Class', 'Batch')]

# Set colors for later plotting
Class <- factor(y.peptide$targets$Class)
Class.color <- Class
levels(Class.color) <- hcl.colors(nlevels(Class), palette = "cividis")

Batch <- factor(y.peptide$targets$Batch)
Batch.color <- Batch
levels(Batch.color) <- hcl.colors(nlevels(Batch), palette = "Set 2")

y.peptide <- filterNonProteotypicPeptides(y.peptide)
y.peptide <- filterCompoundProteins(y.peptide)
y.peptide <- filterSingletonPeptides(y.peptide, min.n.peptides = 2)

# Remove contaminant proteins from the genes dataframe
y.peptide$genes <- y.peptide$genes %>% filter(!str_detect(Protein.Group, "Cont_"))

# Match peptides in the matrix to only those remaining in the dataframe
y.peptide$E <- y.peptide$E[rownames(y.peptide$genes), ]

dpcfit <- dpcON(y.peptide, robust=TRUE)

# Print the intercept and slope of our DPC
dpcfit$dpc


y.protein <- dpcQuant(y.peptide, "Protein.Names", dpc=dpcfit)


```

## Differential Expression Analysis (DEA)

Now it is time to analyse our data and investigate proteins that are differentially expressed between our two experimental groups (control vs active Crohn’s disease). We will make use of the `limma` package and closely related `limpa` functions specifically designed for proteomics data. 

First, we will **fit a linear model** that includes both Class and Batch effects.

```{r}
# Set the control group as our reference
Class <- relevel(Class, ref = "Ctrl")
## We also need to reset our colors to match the correct labels
Class.color <- Class
levels(Class.color) <- hcl.colors(nlevels(Class), palette = "cividis")

# Fit a linear model
design <- model.matrix(~Class + Batch)
design
```

We can see both Class and Batch accounted for in our design matrix. 

Now we run the `limpa` **differential expression analysis** function `dpcDE()` to identify the top differentially expressed proteins in our dataset.

```{r}
fit <- dpcDE(y.protein, design, plot = TRUE)
fit <- eBayes(fit)

# Save our DEA results as a dataframe
results <- topTable(fit, coef = 2, number = Inf)
head(results, n=10)
```

:::discussion

# A closer look at the results

Our output table is ordered by B statistic. Use `View(results)` 

:::

## Visualise differentially expressed proteins

### Expression values and standard error by sample

We can visualise the log2 expression values for individual proteins, together with standard errors.

```{r}
plotProtein(y.protein, "PRTN3_HUMAN", col = as.character(Class.color))
legend('topleft', legend = levels(Class), fill = levels(Class.color))

plotProtein(y.protein, "S10A9_HUMAN", col = as.character(Class.color))
legend('topleft', legend = levels(Class), fill = levels(Class.color))
```

:::::challenge

This simple function plots the protein values from our expression matrix, which we did not directly edit to mitigate batch effects. Try running the `limma` function to remove batch effects from our dataset and see how this changes results.

:::solution

```{r}
y.protein.rbe <- y.protein
y.protein.rbe$E <- removeBatchEffect(y.protein,batch = y.protein$targets$Batch, group=y.protein$targets$Class)

plotProtein(y.protein.rbe, "S10A9_HUMAN", col = as.character(Class.color))
legend('topleft', legend = levels(Class), fill = levels(Class.color))

```

:::
:::::

### Heatmap

We can filter our results to the up to 50 top most significant differentially expressed proteins, then visualise via a clustered heatmap.

```{r}
# Filter for the up to 50 most significant results
sig_proteins <- results %>%
  filter(adj.P.Val < 0.05) %>% # filter by p value
  top_n(50, wt = abs(logFC)) %>% # filter by absolute log fold change
  pull(Protein.Names)
expr_matrix <- y.protein$E[sig_proteins, ]

# Scale the data and visualise via heatmap
scaled_expr <- t(scale(t(expr_matrix)))

pheatmap(scaled_expr,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         annotation_col = y.protein$targets,
         clustering_method = 'ward.D')
```

:::challenge

How would you interpret this heatmap?

:::::

::::::::::::::::::::::::::::::::::::: instructor

## Heatmap interpretation

As expected, our samples cluster into their respective classes. 

The three proteins with decreased expression in the aCD samples are clearly distinguishable from those with increased expression in the aCD samples.

There are no apparent batch effects confounding these results.

::::::::::::::::::::::::::::::::::::::::::::::::

### Volcano plot

We can also visualise our results as a volcano plot using `EnhancedVolcano()`.

```{r message=FALSE,warning=FALSE}
EnhancedVolcano(results,
                lab = results$Genes,
                x = 'logFC',
                y = 'adj.P.Val',
                pCutoff = 0.05,
                FCcutoff = 1.0,
                pointSize = 2.0,
                labSize = 3.5,
                drawConnectors = TRUE,
                maxoverlapsConnectors = Inf,
                lengthConnectors = unit(0.001, 'npc'),
                typeConnectors ='closed',
                ylim = c(0, 5),
                xlim = c(-5, 5))
```


::::::::::::::::::::::::::::::::::::: keypoints 

- We can fit a linear model to our data and conduct differential expression analysis using `limpa` and `limma` functions.
- There are many different ways to visualise differentially expressed proteins.

::::::::::::::::::::::::::::::::::::::::::::::::

