---
title: 'Network and Enrichment Analysis'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- What is DE analysis

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- To do DE analysis

::::::::::::::::::::::::::::::::::::::::::::::::

## Network and Enrichment Analysis

### STRING Protein–Protein Interaction Network

Note: this part does not seem to be working

```{r, eval=FALSE}

string_db <- STRINGdb$new(version = "11.5", species = 9606, score_threshold = 400)
mapped <- string_db$map(results, "Protein.Names", removeUnmappedRows = TRUE)
de_proteins <- mapped %>% filter(adj.P.Val < 0.05)

# Plot network of significant proteins
string_db$plot_network(de_proteins$STRING_id)

```

### Functional Enrichment (GO and KEGG)

```{r}

de_proteins <- results %>% filter(adj.P.Val < 0.05)

# Convert UniProt IDs to Entrez IDs for enrichment
converted <- bitr(de_proteins$Protein.Group,
                  fromType = "UNIPROT",
                  toType = "ENTREZID",
                  OrgDb = org.Hs.eg.db)

# GO Biological Process enrichment
ego <- enrichGO(gene = converted$ENTREZID,
                OrgDb = org.Hs.eg.db,
                ont = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                readable = TRUE)

dotplot(ego, showCategory = 10)

# KEGG pathway enrichment
ekegg <- enrichKEGG(gene = converted$ENTREZID,
                    organism = 'hsa',
                    pvalueCutoff = 0.05)

dotplot(ekegg, showCategory = 10)


```


## Potential next steps

* Compare coefficients between classes as an exercise.

* Visualize selected proteins of interest across experimental groups.

* Extend enrichment analyses using Reactome or GSEA approaches.

* Integrate sample metadata for clinical covariates or longitudinal modeling.




::::::::::::::::::::::::::::::::::::: keypoints 

- DEA

::::::::::::::::::::::::::::::::::::::::::::::::

